Towards Practical First-Order Model Counting

### 2024-04-26 doc/paper.tex
Most of the introduction

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item 9 pages including appendices but not acknowledgements
    \item Add some papers mentioned in:
          \begin{itemize}
            \item very recent work
            \item my previous paper, including:
                  \begin{itemize}
                    \item other liftable fragments
                    \item some more theory papers, e.g., LICS 2018
                  \end{itemize}
          \end{itemize}
    \item introduce \Cranetwo{}
    \item (maybe) introduce lifted/liftable
  \end{itemize}
}

\paragraph{Papers To Cite}
\begin{itemize}
  \item overviews
        \begin{itemize}
          \item lifted probabilistic inference~\cite{DBLP:conf/ecai/Kersting12}
          \item recent overview paper~\cite{DBLP:conf/ijcai/Kuzelka23}
        \end{itemize}
  \item Alternative definition~\cite{DBLP:journals/cacm/GogateD16}
  \item relevant theoretical work~\cite{DBLP:conf/aaai/MalhotraS22}
  \item original domain recursion~\cite{DBLP:conf/nips/Broeck11}
  \item algorithms
        \begin{itemize}
          \item \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
          \item \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}
          \item \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}
          \item \textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16} (similarly to us
                compiles to C++ code, but (probably) doesn't work on as many
                formulas)
          \item approximate~\cite{DBLP:conf/ijcai/BremenK20}
          \item for Markov logic networks~\cite{DBLP:journals/ml/RichardsonD06}
                \begin{itemize}
                  \item \textsc{Magician}~\cite{DBLP:conf/aaai/VenugopalSG15}
                  \item \textsc{Tuffy}~\cite{DBLP:journals/pvldb/NiuRDS11}
                  \item \textsc{Alchemy}~\cite{DBLP:journals/cacm/GogateD16}
                        (same as the alternative definition)
                \end{itemize}
        \end{itemize}
  \item complexity
        \begin{itemize}
          \item liftability~\cite{DBLP:conf/starai/JaegerB12}
          \item hardness for three variables~\cite{DBLP:conf/pods/BeameBGS15}
          \item liftable fragments
                \begin{itemize}
                  \item \Ctwo~\cite{DBLP:journals/jair/Kuzelka21}
                  \item tree axioms~\cite{DBLP:journals/ai/BremenK23}
                  \item linear order axioms~\cite{DBLP:conf/aaai/TothK23}
                  \item some liftable fragments~\cite{DBLP:conf/nips/KazemiKBP16}
                \end{itemize}
        \end{itemize}
  \item applications
        \begin{itemize}
          \item extensions to
                sampling~\cite{DBLP:conf/aaai/WangB0K22,DBLP:conf/lics/WangP0K23}
          \item discovery of combinatorial
                sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23}
          \item conjecturing recurrence
                relations~\cite{DBLP:conf/ilp/BarvinekB0ZK21}
          \item probabilistic logic
                programming~\cite{DBLP:journals/ijar/RiguzziBZCL17} (WFOMC was
                shown to be supreme)
          \item probabilistic databases~\cite{DBLP:journals/debu/GribkoffSB14}
        \end{itemize}
  \item lifted inference elsewhere
        \begin{itemize}
          \item constraint satisfaction~\cite{DBLP:journals/jair/TotisDRK23}
          \item answer set programming~\cite{DBLP:journals/ijar/AzzoliniR23}
        \end{itemize}
\end{itemize}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item By pointing to the outline in \cref{fig:overview}, make forward
          references to the sections of the paper.
    \item Somewhere: mention that we're primarily concerned with computing
          unweighted model counts, but everything trivially extends to the
          weighted setting as well.
    \item Introduce greedy search and breadth-first search (BFS), together with
          \Cranegreedy{} and \Cranebfs{}. The knowledge compilation part of both
          \textsc{Crane} and \Cranetwo{} can be executed using either greedy
          (similar to \textsc{ForcLift}) or breadth-first search. We use both in
          our experiments, denoting them as \Cranegreedy{} and \Cranebfs{},
          respectively.
    \item Introduce the WFOMC acronym.
    \item Introduce first-order knowledge compilation as a class of WFOMC
          algorithms, defining/mentioning compilation rules, FCG, etc.
  \end{itemize}
}

\paragraph{Contributions}
\begin{itemize}
  \item Completing the definitions of recursive functions by:
  \begin{itemize}
    \item identifying a sufficient set of base cases (\cref{sec:identifying})
    \item constructing formulas that correspond to these base cases
          (\cref{sec:simplifying})
    \item and recursing on these subproblems
  \end{itemize}
  \item Compiling these function definitions into a C++ program that can be
        executed independently for any domain size values \cref{sec:cpp}
        \begin{itemize}
          \item including support for infinite precision arithmetic via GNU
                Multiple Precision Arithmetic Library
        \end{itemize}
  \item Experiments comparing \Cranetwo{} with the main alternative approach
        demonstrate the ability of \Cranetwo{} to scale to domain sizes\dots
\end{itemize}

\Cref{sec:identifying,sec:cpp} deal with algebraic constructs whereas
\cref{sec:simplifying} deals with logic.

@@

\begin{remark}
  In the case of \textsc{ForcLift} and its extensions, support for a formula as
  valid input does not imply that the algorithm can compile the formula into a
  circuit or graph suitable for lifted model counting. However, it is known that
  \textsc{ForcLift} compilation is guaranteed to succeed on any \FO{} formula
  without constants and with at most two
  variables~\cite{DBLP:conf/nips/Broeck11}.
\end{remark}

@@

\Cref{alg:compilewithbasecases} outlines our overall approach for compiling a
formula into a set of equation that include the required base cases. In short,
we first use the knowledge compilation algorithm of the original
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23} to compile the formula into:
\begin{enumerate*}[label=(\roman*)]
  \item set $\mathcal{E}$ of equations,
  \item map $\mathcal{F}$ from function names to formulas, and
  \item map $\mathcal{D}$ from function names and argument indices to domains.
\end{enumerate*}
After some algebraic simplification, $\mathcal{E}$ is passed to the
\FindBaseCases procedure (described in \cref{sec:identifying}) that returns a
set of base cases that we need to find solutions for. For each base case
$f(\mathbf{x})$, we identify the formula associated with $f$ and simplify it
using the \Propagate procedure (described in \cref{sec:simplifying}). The
algorithm then recurses on these simplified formulas and adds the resulting base
case equations to $\mathcal{E}$. \Cref{example:overall} explains
\cref{alg:compilewithbasecases} in more detail.

\begin{example}\label{example:overall}
  Consider the following formula (previously examined by Dilkas and
  Belle~\shortcite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ to be a
  bijection between two sets $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z\\
      \forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z.
    \end{gathered}
  \]
  In particular, we examine the first solution that \Cranebfs{} returns for this
  formula.

@@

\todo[inline]{Can I turn the above into a theorem?}

@@

\todo[inline]{A concluding sentence that introduces the remaining subsections?}

@@

\section{Smoothing}

Smoothness originates in propositional knowledge compilation where it is defined
as the property that, for every disjunction node, all disjuncts (as subtrees of
the circuit) contain the same atoms~\cite{darwiche2001tractable}. Van den Broeck
et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} generalise smoothness to
first-order logic, adding set-disjunction and inclusion-exclusion nodes
alongside disjunction.

@@

In the rest of this section, we extend the smoothing algorithm of Van den Broeck
et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} even further.
\Cref{sec:smoothingbase} describes the role smoothing plays in the
base-case-finding algorithm from \cref{sec:main}. Then, \cref{sec:smoothingfcg1}
shows how to adapt smoothing to the compilation rules introduced by Dilkas and
Belle~\shortcite{DBLP:conf/kr/DilkasB23}.

@@

\section{Experimental Evaluation}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Later on:
          \begin{itemize}
            \item Eliminate (e.g., Emacs) warnings.
            \item must run it by all 3 other coauthors
            \item re-check submission instructions and formatting guidelines
            \item cite the \textsc{Crane} paper (and other algorithms' papers)
                  where necessary
          \end{itemize}
  \end{itemize}
}

>>>>

\crefalias{enumi}{step}
\crefname{step}{Step}{Steps}
\creflabelformat{step}{#2\textup{#1}#3}

@@

% 1. What is the problem?

\emph{First-order model counting} is the problem of counting the number of
models of a sentence in first-order logic over some given domain(s). The
(symmetric) weighted variant of the problem (WFOMC) asks for the sum of model
weights instead, where the weight of a model is a product over predicate
weights~\cite{DBLP:conf/ijcai/BroeckTMDR11}. WFOMC is a prominent approach to
the broader problem known as \emph{lifted (probabilistic) inference} that aims
to compute probabilities more efficiently by utlilising symmetries inherent in
the description of the problem~\cite{DBLP:conf/ecai/Kersting12}.

% 2. Why is it interesting and important?

Lifted inference is an active area of research with recent work in domains such
as constraint satisfaction problems~\cite{DBLP:journals/jair/TotisDRK23} and
probabilistic answer set programming~\cite{DBLP:journals/ijar/AzzoliniR23}.
WFOMC itself has been used to perform inference on probabilistic
databases~\cite{DBLP:journals/debu/GribkoffSB14} and probabilistic logic
programs~\cite{DBLP:journals/ijar/RiguzziBZCL17}. By iteratively considering
domains of increasing sizes, the model count of a formula can be interpreted as
an (integer) sequence. WFOMC algorithms have been used to discover new
sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23} and
conjecture~\cite{DBLP:conf/ilp/BarvinekB0ZK21} and
construct~\cite{DBLP:conf/kr/DilkasB23} recurrence relations and other recursive
constructs that describe these sequences. Recently, WFOMC algorithms have also
been extended and applied to perform
\emph{sampling}~\cite{DBLP:conf/aaai/WangB0K22,DBLP:conf/lics/WangP0K23}.

% 3. Why is it hard? (E.g., why do naive approaches fail?)

The complexity of WFOMC is typically characterised in terms of \emph{data
  complexity}, i.e., fixing the formula and asking whether there exists an
algorithm that can compute the WFOMC in time polynomial in the domain size(s).
If such an algorithm exists, the formula is called
\emph{liftable}~\cite{DBLP:conf/starai/JaegerB12}. Beame et
al.~\shortcite{DBLP:conf/pods/BeameBGS15} showed that there exists an unliftable
formula with three variables. It is also known that all formulas with up to two
variables are liftable~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}.
This liftable fragment of formulas with two variables has been extended with
various axioms~\cite{DBLP:conf/aaai/TothK23,DBLP:journals/ai/BremenK23},
counting quantifiers~\cite{DBLP:journals/jair/Kuzelka21}, and in other
ways~\cite{DBLP:conf/nips/KazemiKBP16}.

% 4. Why hasn't it been solved before? (Or, what's wrong with previous proposed
% solutions? How does mine differ?)

There is a variety of WFOMC algorithms with different underlying principles.
Perhaps the most prominent class of WFOMC algorithms is based on
\emph{first-order knowledge compilation}. Here, by iteratively applying
\emph{compilation rules}, the formula is compiled to a representation (such as a
circuit of graph), which can then be used to compute the WFOMC for any
combination of domain sizes (or weights). Algorithms in this class include
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its recent extension
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}. The former compiles formulas into
circuits whereas the latter compiles them first to \emph{first-order
  computational graphs} (FCGs) and then to (algebraic) equations. Another major
WFOMC algorithm \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21} is based on
cell enumeration. Other algorithms employ local
search~\cite{DBLP:journals/pvldb/NiuRDS11}, junction
trees~\cite{DBLP:conf/aaai/VenugopalSG15}, Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}, and anytime approximation via
upper/lower bound construction~\cite{DBLP:conf/ijcai/BremenK20}.

% 5. What are the key components of my approach and results? Also include any
% specific limitations.

The recently-proposed \textsc{Crane} algorithm, while able to handle formulas
beyond the reach of \textsc{ForcLift}, was somewhat incomplete in that it could
only construct functions definitions, which would then need to be evaluated to
compute the WFOMC.\@ Moreover, recursive functions were presented without base
cases. In this work, we present \Cranetwo{}---an extension of \textsc{Crane}
that fixes these two weaknesses.

\todo[inline,caption={}]{
  \begin{itemize}
  \item Finish the paragraph above.
  \item mention that \textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16} compiles to C++
  \end{itemize}
}

@@

\Cref{fig:overview} outlines the workflow of the new algorithm \Cranetwo{}. In
\cref{sec:main}, we describe how \CompileWithBaseCases finds the base cases for
recursive functions by:
\begin{enumerate*}[label=(\roman*)]
  \item identifying a sufficient set of base cases for each function,
  \item constructing formulas that correspond to these base cases,
  and\label[step]{step2}
  \item recursing on these new formulas.
\end{enumerate*}
Then, \cref{sec:smoothing} describes post-processing techniques for FCGs and the
formulas from \cref{step2} above necessary to preserve the correct model count.
Next, \cref{sec:cpp} explains how function definitions that encode a solution to
a WFOMC problem are then compiled to a C++ program. Finally,
\cref{sec:experiments} reports on two experiments that compare \Cranetwo{} with
other state-of-the-art WFOMC algorithms.

@@

\begin{remark}
  In the case of \textsc{ForcLift} and its extensions, support for a formula as
  valid input does not imply that the algorithm can compile the formula into a
  circuit or graph suitable for lifted model counting. However, it is known that
  \textsc{ForcLift} compilation is guaranteed to succeed on any \FO{} formula
  without constants and with at most two
  variables~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}.
\end{remark}

@@

\Cref{alg:compilewithbasecases} outlines our overall approach for compiling a
formula into a set of equation that include the required base cases. In short,
we first use the knowledge compilation algorithm of the original \textsc{Crane}
to compile the formula into:
\begin{enumerate*}[label=(\roman*)]
  \item set $\mathcal{E}$ of equations,
  \item map $\mathcal{F}$ from function names to formulas, and
  \item map $\mathcal{D}$ from function names and argument indices to domains.
\end{enumerate*}
After some algebraic simplification, $\mathcal{E}$ is passed to the
\FindBaseCases procedure (described in \cref{sec:identifying}) that returns a
set of base cases that we need to find solutions for. For each base case
$f(\mathbf{x})$, we identify the formula associated with $f$ and simplify it
using the \Propagate procedure (described in \cref{sec:simplifying}). The
algorithm then recurses on these simplified formulas and adds the resulting base
case equations to $\mathcal{E}$. \Cref{example:overall} explains
\cref{alg:compilewithbasecases} in more detail.

\begin{example}\label{example:overall}
  Consider the following formula (previously examined by Dilkas and
  Belle~\shortcite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ to be a
  bijection between two sets $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z\\
      \forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z.
    \end{gathered}
  \]
  In particular, we examine the first solution that \Cranetwo{} returns for this
  formula.

@@

@@

@@

\section{Smoothing}\label{sec:smoothing}

\emph{Smoothness} originates in propositional knowledge compilation where it is
defined as the property that, for every disjunction node, all disjuncts (as
subtrees of the circuit) contain the same atoms~\cite{darwiche2001tractable}.
Van den Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} generalise
smoothness to first-order logic, adding set-disjunction and inclusion-exclusion
nodes alongside disjunction.

@@

The rest of this section presents an extension to the smoothing algorithm of Van
den Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}.
\Cref{sec:smoothingbase} describes the role smoothing plays in the
base-case-finding algorithm from \cref{sec:main}. Then, \cref{sec:smoothingfcg1}
shows how to adapt smoothing to the compilation rules introduced by Dilkas and
Belle~\shortcite{DBLP:conf/kr/DilkasB23}.

@@

\section{Experimental Evaluation}\label{sec:experiments}

\todo[inline]{Introduce greedy search and breadth-first search (BFS), together
  with \Cranegreedy{} and \Cranebfs{}. The knowledge compilation part of both
  \textsc{Crane} and \Cranetwo{} can be executed using either greedy (similar to
  \textsc{ForcLift}) or breadth-first search. We use both in our experiments,
  denoting them as \Cranegreedy{} and \Cranebfs{}, respectively.}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Later on:
          \begin{itemize}
            \item Eliminate (e.g., Emacs) warnings.
            \item must run it by all 3 other coauthors
            \item re-check submission instructions and formatting guidelines
          \end{itemize}
  \end{itemize}
}
